Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2020-05-05 analysis/paper/paper.Rmd
add site locations CSV for updates

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human]. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- generally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) --> 
<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths. In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes. The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance]. The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

<!-- the last sentense is a bit long-->
<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->

The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca. 36,000 cal BP) [@yamaoka2012use; @izuho201421]. The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421]. These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421].

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that --> 
<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals. Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP). Bae’s earlier stage is started from Koreyri site dated to 27 ka. The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points. Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin]. An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####]. The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades. Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip. These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.

@@

>>>>

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations). Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations). To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- sgenerally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation). The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain. <!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate? -->

The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools (citations). New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools. <!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. <!-- need to expand, mention some key site, blades appear when? what is the debate? -->

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. <!-- need to expand, mention some key site & details. what is the debate?  -->

@@

write_csv(here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
